Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}

### 2018-01-05 drafts/elife/draft.tex
adding updated figures.

\begin{figure}[t!]
\centering%
\includegraphics[width=6in]{figures/v2/figure1} 
\caption{
  \textit{A probabilistic framework for combining
  information across multiple worms in order to learn a shared
  dynamical system for \celegans.}
  \textbf{a.}~The data consists of observed neural activity collected
  from many different worms.  In each recording we observe a different
  subset of labeled neurons in the head ganglia.
  \textbf{b.}~We model the observed activity as a linear function of
  an underlying, continuous, and typically low-dimensional latent state,
  here shown as a two-dimensional signal for each recording.
  \textbf{c.}~The dynamics of these continuous states are governed by
  a co-evolving discrete state.
  \textbf{d.}~These discrete states index into a library of linear
  dynamics functions for each worm, here shown as maps from~$\reals^2$ to~$\reals^2$.
  Each dynamics function is selectively used in different regions of
  continuous latent space, as shown by the different intensities of
  the vector field.
  \textbf{e.}~To increase robustness to model misspecification, we
  use a multivariate-t distribution for the noise in the continuous
  state updates.
  \textbf{f.}~Each worm has its own dynamics library, but they are
  constrained to be close to the shared, canonical library.  This
  combines information to learn shared structure while also allowing
  for worm-to-worm variability.
  \textbf{g.} Likewise, we allow each worm to have its own observation
  variances to account for individual differences that may arise from,
  for example, variable levels of fluorescent protein expression.
  Missing neurons are denoted by gray boxes. 
}
\label{fig:model}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v2/figure2part1} 
\caption{ \textit{Hierarchical state space models smooth noisy recordings and
  make predictions about activity of unseen neurons.}
  \textbf{a.}~The observed neural activity for the
  five worms of the \citet{kato2015global} dataset along with the
  smoothed activity under the hierarchical model.  Only a subset of neurons
  are labeled in each worm; the rest are treated as missing data.
  The model uses correlations learned across the population to predict the activity
  of unobserved neurons.
  The neurons have been clustered based on similarity, as shown in Fig.~\ref{fig:clustering}.
  \textbf{b.} The observed activity (green lines) and smoothed activity (black lines)
  for 3 minutes of activity from worm 5, as denoted by the dotted black box in panel a.
  Neurons are in the same order as panels a and c.
  \textbf{c.} Zooming in on 30 seconds of activity denoted by the box in panel b.
}
\label{fig:smoothing}
\end{figure}

The parameters of the learned model offer novel insights into the
structure of neural activity in the head ganglia that can only be
obtained by pooling information across many partial recordings.  With
this aggregated information, we can smooth noisy observations and
predict the activity of unobserved neurons. \citet{kato2015global}
used a smoothing procedure to estimate temporal principal components.
By contrast, we fit our model directly to the first-order differences
in calcium fluorescence; our only preprocessing step is to correct for
photo-bleaching. Fig~\ref{fig:smoothing}a show our results for the 
five worms in this dataset. We see that the model smooths out
noise in the observed activity and, moreover, makes predictions about
the activity of neurons that could not be identified in this worm.

Fig.~\ref{fig:smoothing}b and shows the segment of neural activity
denoted by the dotted black box in a. The neurons are clustered based
on similarity, as discussed in the next section.  Here, the green lines
show the observed activity and the black lines show the smoothed activity.
Neurons that were not identified in this worm do not have observed activity,
but the model still makes predictions about their activity. Fig.~\ref{fig:smoothing}c
zooms in further to show a 30 sec window of activity. This shows the
degree of noise in the raw data.

Even though each individual recording contains only a subset of neurons,
the hierarchical model can still estimate the dimensionality of
neural activity in the entire head ganglia.  We reserve the last 20\% of
the data (3.6 min) for testing and estimate the dimensionality
of neural activity based on the marginal log likelihood the model assigns
to this held-out data. We find that the test likelihood peaks with a
ten-dimensional latent state (Fig.~\ref{fig:supp_emissions}c).  Moreover, the likelihood is substantially
higher in the model that includes per-worm observation variances
compared to the model with shared variances for all worms. Fig.~\ref{fig:supp_emissions}d
shows the inferred variance for each worm in the \citet{kato2015global}
dataset.  Neurons that were not identified are marked with a gray box.
This shows the extent of the missing data and the need for hierarchical
modeling.\todo{make supplement figure}

\subsection*{Predictions aid in neural identification}

\todo[inline]{Make a figure for this result}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v2/figure2part2} 
\caption{ \textit{The mapping from continuous latent states reveals clusters of
    functionally similar neurons.}
  We can view the linear mapping from latent states to observations
  as either a matrix (\textbf{a.}) or as an embedding of neurons (\textbf{b.}).
  \textbf{c.}~Viewed as an embedding, we see that neurons cluster into
  correlated groups with similar tuning to the continuous latent state.
  \textbf{d.}~The embedding similarity reflects correlations between
  neurons, but since it is learned as part of a hierarchical model, it
  predicts similarities for all pairs of neurons, even those that are
  never simultaneously recorded.  These pairs are denoted by gray boxes
  in the empirical correlation matrix.
  \textbf{e.}~Neurons in the same cluster are not necessarily close
  to one another in space, though we do see that, on average, neurons in cluster 2
  are more anterior whereas neurons in clusters 5 and 6 are more posterior.}
\label{fig:clustering}
\end{figure}

% --------- neural identification ----------------
% The predicted activity is based on the correlations between neurons
% learned from many recordings.  These predictions are a useful aid for
% identifying unlabeled neurons.  For example, we see that
% \textsf{SMBDL} was not found in the first worm
% (Fig.~\ref{fig:emissions}a). If the activity of one of the unlabeled
% neurons closely matches the model's predictions for \textsf{SMBDL},
% then this provides some evidence in favor of assigning it this
% label. We tested this idea by withholding the labels of some
% confidently-identified neurons and comparing their activity to the
% model's predicted activity for all unlabeled neurons.  We measured the
% similarity between the actual and predicted activity and found that
% correct label, or ``candidate,'' tends to be much more similar than
% other incorrect candidates, as summarized in
% Fig.~\ref{fig:emissions}b.  Combined with other side-information, like
% the location of the neuron, this provides a powerful aid for identifying
% neurons.  The more neurons that we can identify, the better we will be
% able to learn about the dynamics of neural activity.

The linear mapping from latent states to observations reveals clusters
of similarly tuned neurons.  The mapping maybe viewed as either an $N$
(neurons) by $D$ (latent dimensions) matrix or as a collection of~$N$
vectors in $D$-dimensional space, i.e. as an embedding of neurons.
These two views are shown in Fig.~\ref{fig:emissions}d. We note that
many vectors point in similar directions, which implies that there are
groups of correlated neurons.  We clustered the neurons based on the
cosine similarity between their embedding vectors and found five clusters
of neurons shown in Fig.~\ref{fig:emissions}f (order of neurons is the same
as in Fig.~\ref{fig:emissions}a). To make sense of these clusters, we then
looked at their spatial distribution along the head-to-tail axis of the
worm's body.  We found that clusters 2-4 correspond to activity closer to
the head and working backward toward the tail.  The last and largest cluster
contains many neurons in the nerve ring (location 0.15) \todo{check} as well
as some posterior neurons. \todo[inline]{discuss functional significance of clusters.}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v2/figure3part1} 
\caption{ \textit{The hierarchical SLDS automatically segments neural
  activity based on its dynamics.}
  \textbf{a.}~Three dimensions of the inferred continuous latent states
  for each of the five worms, color coded by the corresponding discrete latent
  state.
  \textbf{b.}~We select the number of discrete states and justify the
  hierarchcial, robust, and recurrent extensions by measuring the marginal likelihood
  of held-out test data. We find that the hierarchical, robust, recurrent model
  with~$K=8$ states performs best.
  \textbf{c.}~The inferred segmentations align with the manual segmentations
  of~\citet{kato2015global}.
  \textbf{d.}~Moreover, the inferred states are in close correspondence with
  the manually labeled states, which were shown to indicate different behaviors.
}
\label{fig:syllables}
\end{figure}

Having learned a mapping from latent state space to observed activity,
we can now reason about the dynamics of these continuous latent
states. Fig.~\ref{fig:syllables}a plots the trajectory of the first
worm's neural activity in the first three dimensions of the continuous
state space, color coded by the discrete regime inferred under the
SLDS.  We see that different regimes correspond to distinct loops
through latent space.  Compare this with Fig. 4b
of~\citet{kato2015global}---they achieved a similar result by manually
segmenting smoothed trajectories in the space spanned by the first
three principal components.  In contrast, our probabilistic framework
is automatic, operates on ten-dimensional latent states, and requires
no additional smoothing.

Moreover, we can use the likelihood of held-out test data to determine
the number of discrete regimes and assess the value of the robust and
recurrent model extensions, as shown in Fig.~\ref{fig:syllables}b.
While the non-robust models with Gaussian dynamics noise achieve
higher training likelihood, the robust models generalize better to
test data.  Moreover, the recurrent models provide further improvement
over the standard Markovian models.  We find that test likelihood is
maximized with eight latent states, which matches the number manually
chosen by~\citet{kato2015global}.

Not only does the SLDS find the same number of states, it also finds a
close correspondence between the manually labeled and inferred states.
Fig.~\ref{fig:syllables}c shows ten-dimensional latent states and the
manual and inferred segmentations for a 5.5 minute section of data.
The manual and inferred states often change at similar times. We also
see instances where there appears to be a change in continuous state
dynamics that are identified by the SLDS but missed in the manual
segmentation (e.g. around $t=700$ seconds).  Fig.~\ref{fig:syllables}d
quantifies the overlap in states for the first worm and for all worms.
Many inferred states are in one-to-one correspondence with manually
labeled states, and the remainder tend to be split between two manual
states.  This confirms that the SLDS is a useful tool for automatically
segmenting brain data.

\subsection*{Quantifying the dynamics of neural activity in each discrete state}
\begin{figure}[h!]
\centering
\includegraphics[width=6in]{figures/v2/figure3part2} 
\caption{ \textit{Each state is characterized by simple linear dynamics.}
    \textbf{a.}~Each discrete state is associated with a canonical set of linear
  dynamics, which can be viewed as mappings from~$\reals^{10}$ to~$\reals^{10}$.
  We show the mappings for the first three dimensions here, with vectors placed
  at locations where each state was used.
  \textbf{b.}~We simulate these linear dynamics from a variety of starting
  locations and compute the time course of resulting activations for each
  cluster of neurons.
  \textbf{c.}~The eigenspectra of the dynamics matrices highlight the rotational
  vs. contractive nature of the states and illustrate worm-to-worm variability. 
}
\label{fig:dynamics}
\end{figure}

Beyond simply automating a manual process, the SLDS also provides a
quantitative description of the different states.  In particular, each
state is associated with a set of linear dynamics.
Fig.~\ref{fig:dynamics}a shows the canonical dynamics associated with
each of the eight discrete states. The arrows in these vector fields
show where the continuous state is mapped to under the corresponding
discrete state (for the first three latent dimensions).  Arrows are
only drawn at those locations where the discrete state was actually
employed.  We see that these dynamics capture the rotational patterns
of the various loops.

We use these dynamics to simulate how brain activity would evolve
under the different states. Fig.~\ref{fig:dynamics}b shows how four
simulations of each state from various starting locations  map 
into activations in the space of neural activity. In particular, we show how they would
activate the six clusters shown in Fig.~\ref{fig:clustering}.  We see
that state 4 corresponds to a pronounced dip followed by a rise in the
activity of cluster 3, alongside a slow ramp in the activity of
clusters 1 and 2.
% Referring back to Fig~\ref{fig:emissions}a, we see
% that cluster 3 consists of \textsf{ALA} and three symmetric pairs of
% neurons (\textsf{ASK}, \textsf{RIV}, and \textsf{SMDV}).  The~\textsf{ASK}
% pair are sensory neurons responsible for initiating local search behavior,
% the \textsf{RIV} pair are polymodal inter- and motor-neurons that specify
% the ventral bias of turns, and the \textsf{SMDV} are motor neurons
% that initiate ventral turns. As state 4 is in close correspondence with
% the ``ventral turn'' state labeled by~\citet{kato2015global}, it makes
% sense that these neurons would be activated.

@@

\subsection*{Recurrent models learn spatially localized states and transition boundaries}

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v2/figure4v2} 
\caption{ \textit{Discrete state transitions show a non-Markovian,
    chain-like structure.}  \textbf{a.}~Counts of inferred transitions
  from one state to another. The states are permuted to emphasize the
  feed-forward structure.  \textbf{b.}~An alternative view of the
  transitions; each state is a node in a graph and edges denote
  high-probability transitions.  This shows transitions between
  forward and reverse crawling linked by dorsal and ventral turns.
  \textbf{c.}~The standard SLDS assumes Markovian transitions, which
  imply a geometric distribution of state durations (dotted black
  lines), but Fig.~\ref{fig:syllables} clearly demonstrates that state
  transitions occur near the origin.  With the exception of state 7,
  which is unstable, the recurrent SLDS incorporates
  this dependency and generates state durations (solid black lines)
  that are a better fit to those inferred from the data (colored
  bars).  }
\label{fig:recurrent}
\end{figure}

The final and most critical component of the dynamical model for
neural activity in the head ganglia is the discrete transition
distribution.  In contrast to standard SLDS, we allow these transition
probabilities to depend on the location in continuous state space (the
\emph{recurrent} dependency) as well as the preceding discrete state
(the \emph{Markovian} dependency).  Fig.~\ref{fig:recurrent}a shows
the contribution of the Markovian term.  Each cell in this matrix
denotes the number of transitions from one state to another. We are
only counting transitions in which the state changes, hence the
diagonal is zero.  We have permuted the states to emphasize the
largely feed-forward structure.

@@

Upon fitting our model, we immediately see a difference in the overall
usage of the eight states (Fig.~\ref{fig:o2}b). The worms in the
lethargus stage predominantly use the blue and yellow states.  Again,
we found a close correspondence between our inferred states and the
manually labeled states of~\citet{nichols2017global}
(Fig.~\ref{fig:o2}c). Here, blue and yellow both map strongly onto the
quiescent (\textsf{Q}) state. This is consistent with the
characterization of the lethargus stage---at this stage, worms exhibit
much less crawling behavior. The pre-lethargus worms execute a broader
variety of behaviors, including reversals (\textsf{Rev.}) and dorsal
and ventral turns  (\textsf{DT} and \textsf{VT}, respectively).

>>>>

\begin{figure}[t!]
\centering%
\includegraphics[width=6in]{figures/v3/figure1} 
\caption{
  \textit{A probabilistic framework for combining
  information across multiple worms in order to learn a shared
  dynamical system for \celegans.}
  \textbf{a.}~The data consists of observed neural activity collected
  from many different worms.  In each recording we observe a different
  subset of labeled neurons in the head ganglia.
  \textbf{b.}~We model the observed activity as a linear function of
  an underlying, continuous, and typically low-dimensional latent state,
  here shown as a two-dimensional signal for each recording.
  \textbf{c.}~The dynamics of these continuous states are governed by
  a co-evolving discrete state.
  \textbf{d.}~These discrete states index into a library of linear
  dynamics functions for each worm, here shown as maps from~$\reals^2$ to~$\reals^2$.
  Each dynamics function is selectively used in different regions of
  continuous latent space, as shown by the different intensities of
  the vector field.
  \textbf{e.}~To increase robustness to model misspecification, we
  use a multivariate-t distribution for the noise in the continuous
  state updates.
  \textbf{f.}~Each worm has its own dynamics library, but they are
  constrained to be close to the shared, canonical library.  This
  combines information to learn shared structure while also allowing
  for worm-to-worm variability.
  \textbf{g.} Likewise, we allow each worm to have its own observation
  variances to account for individual differences that may arise from,
  for example, variable levels of fluorescent protein expression.
  Missing neurons are denoted by gray boxes. 
}
\label{fig:model}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v3/figure2} 
\caption{ \textit{Hierarchical state space models smooth noisy
    recordings and predict the activity of unseen neurons.}  We show
  the neural activity (differenced Ca++ traces) for two of the
  five worms from the \citet{kato2015global} dataset.  Of these five
  worms, worm 1 (left) has the fewest labeled neurons and worm 5
  (right) has the most.  Green lines are the measured activity of the
  labeled neurons; black lines are the smoothed activities predicted
  by the model for the 61 neurons that were labeled in at least one
  worm.  The model uses correlations learned across all recordings to
  predict the activity of unlabeled neurons, like \textsf{SMBDL} in
  worm 1.  The neurons have been clustered based on similarity, as
  discussed in Fig.~\ref{fig:clustering}.  }
\label{fig:smoothing}
\end{figure}

The parameters of the learned model offer novel insights into the
structure of neural activity in the head ganglia that can only be
obtained by pooling information across many partial recordings.  We
fit our model directly to the first-order differences in calcium
fluorescence; our only preprocessing step is to correct for
photo-bleaching. The green lines in Fig.~\ref{fig:smoothing} show a
three minute window of bleaching-corrected, differenced calcium traces
for two of the five worms in the~\citet{kato2015global} dataset. Of
these five worms, Worm 1 has the fewest labeled neurons (24 neurons)
and worm 5 has the most (48 neurons). Across the five worms, 61 unique
neuron labels were identified in at least one worm.

By aggregating information across worms, the model learns correlation
structure that enables it to smooth noisy traces and predict the
activity of unobserved neurons.  The smoothed and predicted activity
is shown as black lines in Fig.~\ref{fig:smoothing}. For example,
even though \textsf{SMDVR} is not labeled in worm 1, the model
learns how its activity covaries with that of other neurons
by leveraging information from other worms, like worm 5, in which
both~\textsf{SMDVR} and~\textsf{SMDVL} are labeled.  Since these
two neurons are highly correlated, the model can make strong predictions
about the activity of~\textsf{SMDVR} in worm 1 on the basis of the
activity of~\textsf{SMDVL}.

The degree to which the smoothed activity matches the observed
activity is governed by the dimensionality of the latent state space.
We estimate this dimensionality with cross validation, reserving the
last 20\% of the data (3.6 min) for testing and measuring the marginal
log likelihood of the held-out data as we increase the latent
dimensionality. We find that the test likelihood peaks with a
10-dimensional latent state (Fig.~\ref{fig:supp_emissions}c).

Moreover, the likelihood is substantially higher in the model that
includes per-worm observation variances compared to the model with
shared variances for all worms. Fig.~\ref{fig:supp_emissions}d shows
the inferred variance for each worm in the \citet{kato2015global}
dataset.  Neurons that were not identified are marked with a gray box.
This shows the extent of the missing data and the need for
hierarchical modeling.\todo{make supplement figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v3/figure3B} 
\caption{ \textit{The mapping from continuous latent states to neural
    activity reveals clusters of functionally similar neurons.}
  \textbf{a.} The emission matrix is a linear map from latent states;
  the rows can be seen as ``tuning vectors'' that specify how each
  neuron responds to changes in the latent state.  These tuning
  vectors cluster into discrete groups (separated by black lines),
  indicating that there are subpopulations of similarly tuned, and
  hence highly correlated, neurons. Neurons are ordered as in
  Fig.~\ref{fig:smoothing}. \textbf{b.}~The pairwise similarity between
  tuning vectors supports this clustering and shows differences
  between subpopulations (clusters have been sorted to make this as
  close to diagonal as possible).  \textbf{c.}~The tuning similarity
  reflects correlations between neurons, but since it is learned as
  part of a hierarchical model, it predicts similarities for all pairs
  of neurons, even pairs that are never simultaneously recorded (gray,
  off-diagonal boxes).  \textbf{d.}~We show the spatial locations of
  neurons from each cluster, as specified by WormAtlas.  Neurons in
  the same cluster are not necessarily close to one another in space,
  though some clusters are spatially localized. Black bars denote
  approximate locations of the anterior (AG), dorsal (DG), ventral
  (VG) and retrovesicular ganglia.}
\label{fig:clustering}
\end{figure}

The linear mapping from latent states to observations reveals clusters
of similarly tuned neurons.  The mapping is given by a matrix with
rows corresponding to neurons and columns corresponding to latent
dimensions, shown in Fig.~\ref{fig:clustering}a.  Each row can be seen
as a \emph{tuning vector} that specifies how the corresponding neuron
responds to each dimension of the latent state.  This matrix reveals
groups of similarly tuned neurons. To capture this, we used k-means to
cluster the neurons based on their tuning vectors and found that 12
clusters adequately captures the diversity of the population.  Using
fewer clusters leads to qualitatively similar groups but omits some of
the finer within-group differences; allowing more clusters eventually
leads to groups of lateral pairs (like \textsf{ASKL} and \textsf{ASKR}
in a.).  The neurons are ordered as in Fig.~\ref{fig:smoothing}, and
clusters are separated by black lines.

Fig.~\ref{fig:clustering}b shows the cosine similarity of tuning
vectors for each pair of neurons, illustrating the differences
and similarities between clusters.  We notice two prominent
features.  First, the block structure of this
matrix lends support to the notion of discrete clusters, indicating
strong similarity between neurons within the same cluster. Second,
we have sorted the clusters in order to make this matrix as
diagonal as possible.  Upon doing so, we see a banded structure
indicating that in addition to the discrete clustering, there is
also a continuum along which these clusters vary, with the first
clusters more similar to each other than to the latter clusters.
Since the diagonal values equal one by definition, these
have been grayed out.

In our model, the emission matrix determines the correlations between
neurons.  As such, we expect the similarity between rows of the
emission matrix to mimic the empirical correlations observed in the
data.  Fig.~\ref{fig:clustering}c shows that this is indeed the case,
as neurons with similar tuning vectors also have higher empirical
correlation coefficients.  However, without a hierarchical model, we
cannot empirically estimate the correlation between pairs of neurons
that were not simultaneously observed.  For this dataset, this leaves
many entries in the empirical correlation matrix missing, as denoted
by the gray off-diagonal entries. A key advantage of our hierarchical
model is its ability to draw inferences in the face of significant
amounts of missing data.

What can explain this clustering?  We first look at the spatial
locations of neurons in each cluster, under the hypothesis that
clusters could be trivially explained by an inability to resolve
differences in activity between nearby neurons.
Fig.~\ref{fig:clustering}d shows the locations of all 61 labeled
neurons, using the positions from WormAtlas\todo{cite} along the
head-tail axis of the worm's body.  The size of the dot indicates the
number of neurons at that location; the bars below denote the
approximate location of the anterior (AG), dorsal (DG), ventral (VG)
and retrovesicular (RVG) ganglia.  Some cluster (e.g. cluster 6) are
spatially localized, but many clusters contain neurons from multiple
head ganglia.  This suggests that location alone cannot explain the
similarity between tuning vectors.

\todo[inline]{Break down clustes by type (sensory/inter/motor
  neuron).}

\todo[inline]{Ask Manuel/Annika to label functional role of neurons.}

\subsection*{Model predictions aid in neural identification}

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v3/figure_id} 
\caption{ \textit{The predicted neural activity is a useful aid for
    neural identification.}  \textbf{a.} The model uses observed and
  labeled activity (green traces) to predict the activity of unlabeled
  neurons, as in Fig.~\ref{fig:smoothing}. \textbf{b.} Alongside the
  labeled activity, we also have traces from many neurons that could
  not be unambiguously labeled. To test whether the model's
  predictions could aid in labeling neurons, we conducted a synthetic
  experiment, withholding the activity of some labeled neurons. Here,
  the red trace is the withheld activity of VB01 and the blue traces
  are the activity of other unlabeled neurons.  Together, these make
  up a pool of candidates for the VB01 label.  \textbf{c.} For each
  candidate, we measured the correlation between its activity and the
  predicted activity of VB01.  \textbf{d.} Repeating this experiment
  50 times, we found that the correct candidates tend to be much more
  correlated with the model's predicted activity than the other,
  incorrect candidates.  Thus, the model's predictions serve as a
  useful aid for labeling neurons.  }
\label{fig:id}
\end{figure}

The predicted activity is based on the correlations between neurons
learned from many recordings.  These predictions should be a useful
aid for laneling neurons whose activity could be measured but whose
identity could not be unambiguously resolved.  To test this
hypothesis, we conducted a synthetic experiment in which we withheld
the labels of ten neurons per worm.  We fit the model to the remaining
neurons and then compared the model's predicted activity to the
activity of the heldout neurons, as well as to the activity of the
unlabeled neurons identified by~\citet{kato2015global}.

Fig.~\ref{fig:id}a shows a schematic of the experiment: though VB01
was labeled in the dataset, we withheld its activity and used the
model to predict it (black trace) based on the observations of the
other, labeled neurons (green traces).  We then compared the model
predictions to the true activity of VB01 (red trace) and to the
activity of other neurons that were identified but not labeled (blue
traces), as shown in Fig.~\ref{fig:id}b.  We computed the correlation
coefficient between these traces and the predicted activity of VB01,
as shown in Fig.~\ref{fig:id}c.
We find that in this case, the model's predictions and the true activity
are highly correlated, even though the true activity was not used
when training the model.  We repeated this experiment for 50 heldout
neurons and found that this pattern holds: the predicted activity is
typically much more correlated with the activity of the correct
heldout trace (red histogram) than with the activity of other unlabeled
neurons (blue histogram).  This suggests that our hierarchical model
may serve an additional role in aiding neural identification in
ambiguous, noisy data.

\todo[inline]{Also discuss simultaneous matching experiments? Possible
  improvements from including more labeled neurons.  Work with Gonzalo
  on learning permutations.  Importance of capturing uncertainty in matching.}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v3/figure4} 
\caption{ \textit{The hierarchical SLDS automatically segments neural
    activity based on its dynamics.}  \textbf{a.}~The top three
  dimensions of the inferred continuous latent states for each of the
  five worms, color coded by the corresponding discrete latent state.
  \textbf{b.}~The inferred segmentations (top) align with the manual
  segmentations (bottom) of~\citet{kato2015global}, though some
  manually labeled states are split into multiple inferred states
  (e.g. forest green state is often split into green and brown) and
  others are combined (e.g. light purple, light blue, and white are
  often combined into pink).  \textbf{c.}~We quantified the overlap
  between the manual and inferred states and found a close
  correspondence.  As~\citet{kato2015global} showed, the manually
  labeled states map onto different behaviors (\textsf{REVSUS}:
  sustained reversal; \textsf{SLOW}: slow forward crawl; \textsf{VT}:
  ventral turn; \textsf{FWD}: forward crawl; \textsf{DT}: dorsal turn;
  \textsf{REV1, REV2}: reversals; \textsf{NOSTATE}:
  undetermined). The correspondence between inferred and manually
  labeled states implies that hierarchical state space models can
  infer behaviorally meaningful states directly from neural activity.}
\label{fig:syllables}
\end{figure}

Having learned a mapping from latent state space to observed activity,
we can now reason about the dynamics of these continuous latent
states. Fig.~\ref{fig:syllables}a plots the trajectory of the first
worm's neural activity in the top three dimensions of the continuous
state space, color coded by the discrete regime inferred under the
SLDS.  These are the three dimensions with the highest percentage of
expained variance. We see that different regimes correspond to
distinct loops through latent space.  Compare this with Fig. 4b
of~\citet{kato2015global}---they achieved a similar result by manually
segmenting smoothed trajectories in the space spanned by the first
three principal components.  In contrast, our probabilistic framework
is automatic, operates on ten-dimensional latent states, and requires
no additional smoothing.

We use the likelihood of held-out test data to determine
the number of discrete regimes and assess the value of the robust and
recurrent model extensions, as shown in Supplemental Fig.~\ref{fig:modelselection}.
While the non-robust models with Gaussian dynamics noise achieve
higher training likelihood, the robust models generalize better to
test data.  Moreover, the recurrent models provide further improvement
over the standard Markovian models.  We find that test likelihood is
maximized with eight latent states, which matches the number manually
chosen by~\citet{kato2015global}.

Not only does the SLDS find the same number of states, it also finds a
close correspondence between the manually labeled and inferred states.
Fig.~\ref{fig:syllables}b shows ten-dimensional latent states and the
manual and inferred segmentations for a 5.5 minute section of data.
The manual and inferred states often change at similar times. We also
see instances where there appears to be a change in continuous state
dynamics that are identified by the SLDS before it is detected in the
manual segmentation (e.g. around 11.8 min).  Fig.~\ref{fig:syllables}c
quantifies the overlap in states for the first worm and for all worms.
Many inferred states are in one-to-one correspondence with manually
labeled states. Some manual states are split into two inferred states
(e.g. ventral turns (\textsf{VT}) are split into green and brown) and
others are combined (e.g. reversals and undetermined states
(\textsf{REV1}, \textsf{REV2}, \textsf{NOSTATE}) map onto pink).
The correspondence between manual and inferred states shows that
the hierarchical SLDS is a useful tool for automatically segmenting
brain data into behaviorally meaningful states.

\subsection*{Quantifying the dynamics of neural activity in each discrete state}
The hierarchical SLDS not only automates the laborious manual
segmentation process, it justifies its segmentation with a
quantitative definition of each inferred state. Each state is
associated with a set of canonical linear dynamics, as shown in
Fig.~\ref{fig:dynamics}a. The arrows in these vector fields show where
the continuous state is mapped to under the corresponding discrete
state (for the top three latent dimensions).  Arrows are only drawn
at those locations where the discrete state was actually employed.

States 1 (blue), 3 (yellow), and 5 (purple) are all characterized
by convergence toward the origin, which corresponds, essentially, to
no change in neural activity. These three states differ in terms of how
quickly they converge to the origin and the amount of rotation as
they do so.  By contrast, states 2 (red), 4 (green), 6 (orange), 7(brown)
and 8 (pink) form parts of rotational loops in latent state space.
As the latent state traces out a loop, it transiently activates
the clusters of neurons that are tuned to the loop direction.
These states are associated with dorsal turns, ventral turns, and
reversals.

We use these dynamics to simulate how brain activity evolves
under the different states. Fig.~\ref{fig:dynamics}b shows how four
simulated trajectories of each state from various starting locations map 
into activations in the space of neural activity. For each discrete state,
we chose the starting locations randomly from the set of locations where
that state was entered in the real data.  We then simulated forward in
time according to that state's dynamics until the simulation predicted
a state change.  We then mapped these four simulations for each state
through the emission matrix and measured the average activation induced
on each of the 12 neural clusters.

These activations lend further evidence in support of the behavioral
labels attributed to each state. For eaxample, state 4 (green) coincides
with the manually labeled ``ventral turn'' state.  Its dynamics produce
a pronounced rise in the activity of cluster 6, which (referring back
to Fig.~\ref{fig:clustering}), consists of the \textsf{RIVL/R} and
\textsf{SMDVL/R} neurons.  The \textsf{RIV} pair are polymodal inter-
and motor-neurons that specify the ventral bias of turns, and the
\textsf{SMDV} are motor neurons that initiate ventral turns.  State 4
dynamics also tend to suppress the activation of cluster 8, which
consists mostly of the \textsf{SMB*} neurons.  As these neurons are
implicated in dorsal turns, this again makes sense. Finally, we see
that states 6 (orange) and 5 (purple) often produce a rise and fall, respectively,
of activation on cluster 8, which together produce a dorsal turn.
These two states do indeed coincide with the manually labeled dorsal
turn state, again confirming the behavioral interpretation of these
dynamical states.

Finally, Fig.~\ref{fig:dynamics}c shows the eigenspectra of the linear
dynamics for each worm and each state.  Imaginary eigenvalues give
rise to rotational dynamics, and the magnitude of the eigenvalues
determines whether the state is stable or not.  If the magnitude is
greater than one (outside the unit disk denoted by the black arc),
the dynamics produce exponential growth in the latent state space; if
the eigenvalues are all less than unit magnitude, the dynamics converge
to a fixed point.  As expected, the rotational states do indeed have
more imaginary eigenvalues, and most states have eigenvalues with
magnitude near one.  Some states, like the brown state, have eigenvalues
larger than one, which produce unstable dynamics, though this is
somewhat balanced by the recurrent extensions discussed in the next
section.  Comparing across rows, we see only minor differences from worm
to worm, substantiating our hierarchical modeling assumptions.

\begin{figure} [b!]
  \caption{ \textit{(Next page.) Discrete states are characterized by
      simple linear dynamics that give rise to interpretable patterns
      of neural activity.}  \textbf{a.}~Each discrete state is
    associated with a canonical set of linear dynamics, which can be
    viewed as mappings from~$\reals^{10}$ to~$\reals^{10}$.  We show
    the mappings for the first three dimensions here, with vectors
    placed at locations where each state was used.  \textbf{b.}~We
    simulated four latent state trajectories with these linear
    dynamics from different starting locations, and then we mapped
    these latent state trajectories through the emission matrix to
    compute a time course of resulting activations for each cluster of
    neurons.  \textbf{c.}~The eigenspectra of the dynamics matrices
    provide insight into the nature of the states: imaginary eigenvalues
    give rise to rotational dynamics and the magnitude determines
    whether the stability of the dynamics. The arc denotes
    the unit magnitude line; eigenvalues to the left are contractive
    and to the right are expansive. Looking across rows, we
    see how the spectra vary from worm to worm. }
\end{figure}
\addtocounter{figure}{-1}

\begin{figure}[h!]
  \centering
  \vspace{-.5in}
  \includegraphics[width=6in]{figures/v3/figure5B}
  \caption{Caption on preceding page.}
  \label{fig:dynamics}
\end{figure}

@@

\subsection*{Recurrent models learn spatially localized states and non-Markovian dynamics}

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v3/figure6} 
\caption{ \textit{Discrete state transitions show a chain-like structure
    with non-Markovian durations.}  \textbf{a.}~Counts of inferred transitions
  from one state to another. The states are permuted to emphasize the
  feed-forward structure.  \textbf{b.}~An alternative view of the
  transitions; each state is a node in a graph and edges denote
  high-probability transitions.  This shows transitions between
  forward and reverse crawling linked by dorsal and ventral turns.
  \textbf{c.}~The standard SLDS assumes Markovian transitions, which
  imply a geometric distribution of state durations (dotted black
  lines), but Fig.~\ref{fig:syllables} suggests that state
  transitions occur near the origin.  With the exception of state 7,
  which is unstable, the recurrent SLDS incorporates
  this dependency and generates state durations (solid black lines)
  that are a better fit to those inferred from the data (colored
  bars).  }
\label{fig:recurrent}
\end{figure}

The final component of our hierarchical state space model is the
discrete transition distribution.  In contrast to standard SLDS, we
allow these transition probabilities to depend on the location in
continuous state space (via the \emph{recurrent} dependency) as well as
the preceding discrete state (the \emph{Markovian} dependency).
Fig.~\ref{fig:recurrent}a shows the contribution of the Markovian
term.  Each cell in this matrix denotes the number of transitions from
one state to another. We are only counting transitions in which the
state changes, hence the diagonal is zero.  We have permuted the
states to emphasize the largely feed-forward structure.

@@

Upon fitting our model, we immediately see a difference in the average
usage of the eight states  as a function of oxygen level(Fig.~\ref{fig:o2}b). The worms in the
lethargus stage predominantly use the blue and red states.  Again,
we found a close correspondence between our inferred states and the
manually labeled states of~\citet{nichols2017global}
(Fig.~\ref{fig:o2}c). Here, blue and red both map strongly onto the
quiescent (\textsf{Q}) state. This is consistent with the
characterization of the lethargus stage---at this stage, worms exhibit
much less crawling behavior. The pre-lethargus worms execute a broader
variety of behaviors, including reversals (\textsf{Rev.}) and dorsal
and ventral turns  (\textsf{DT} and \textsf{VT}, respectively).
